Improved Quasi-Static Methods in \rattlesnake

### 2017-03-09 doc/NEAMS_Report/NEAMS_Report.tex

% input stuff
\lstset{%
%  language=C++,
  showstringspaces=false,
% basicstyle=\scriptsize\ttfamily,
  basicstyle=\small\ttfamily,
  commentstyle=\color{inl@green},
%  keywordstyle=\bfseries,
%  escapeinside=$$
%  framesep=0pt,
%  rulesep=0pt,
  frame=single
}
% Make nice big tildes inside lstlistings instead of
% tiny ones.
\lstset{
    literate={~} {$\sim$}{1}
}

@@

\rattlesnake is a MOOSE-based application developed INL specific to solving radiation transport problems with multiphysics capabilities.  MOOSE is a finite-element based, multiphysics framework that gives the general architecture for the development of physics application like \rattlesnake.   At the heart of \rattlesnake is the action system, which provides a means to consolidate the MOOSE input syntax (which can be quite larger for multigroup transport simulations), so that a user does not have to define every kernel, variable, etc., used in the problem. Rather, the user inputs an equation description 
(e.g., Diffusion, S$_n$, P$_n$, etc.) and a solution method (e.g., SAAF, LS, CFEM, DFEM, etc.), and the action system will incorporate all the necessary physics involved (kernels, boundary conditions, postprocessor, etc.).

@@

The input file for IQS is very similar to the current transient diffusion input file.  The IQS input has a different executioner type and parameters.  The executioner type is simply IQS and input parameters include the number of micro time steps per macro step, the IQS error tolerance, and the initial power.  The \rattlesnake transient action system currently requires a multi-app and transfer to compute and pass the initial $\phi$ and $\keff$, which is present in the transient input deck.  However, IQS also requires an initial evaluation of the adjoint flux, for the weighting function.  So another input file and multi-app transfer was made for the adjoint calculation.  \\

@@

\fig{fig:TWIGL_power} shows the IQS  solution as compared with the Brute Force solution.  These plots show that IQS is consistent in more complex, higher dimensional problems in \rattlesnake.  Finally, \fig{fig:TWIGL_conv} plots the error convergence of IQS and the Brute Force methods.  The curves show the impressive convergence of IQS for the highly transient TWIGL example.  \tbl{tab:TWIGLdt2} and \fig{fig:TWIGL_power_dt2} show the results for time adaptation.  The results show that both IQS methods perform exceptionally well compared to brute force.  It also shows that traditional IQS performed better with large $e_{tol}$, while IQS P-C was better with smaller $e_{tol}$.

@@

The \fig{fig:LRA_plots} plots show the IQS  solution as compared with the Brute Force solution.    These plots show that IQS is consistent with Doppler feedback problems in \rattlesnake.  \fig{fig:lra_conv} plots the error convergence of IQS and the Brute Force methods. \tbl{tab:LRAdt2} shows the time adaptation performance of brute force and IQS P-C.  These results show that IQS, again, performs much better than the brute force method by taking significantly fewer diffusion solves while retaining similar accuracy.

@@

The purpose of this report is to briefly review the improved quasi-static method for neutronic transient simulations, to describe its implementation in \rattlesnake, the radiation transport application based on MOOSE, and to show its effectiveness using several test cases.  
%The derivation of IQS was done with multigroup diffusion, but it was seen in \sct{sect:transport} that application to transport is relatively simple.  
Two IQS implementations are available in \rattlesnake.  The first one is the traditional IQS approach, where the shape equation and PRKE are solved as a coupled nonlinear system with Picard iterations.  The second approach, IQS P-C, is a predictor-corrector linearization, where the flux is solved and then rescaled using the amplitude obtained through the PRKE
(Point Reactor Kinetic Equations).  IQS P-C is a simpler method and is computationally much less expensive to perform. \\

IQS's implementation in \rattlesnake involves a new kernel, an auxiliary kernel, a new executioner derived from the current transient, four postprocessors, as well as alterations to the action system.  The new kernel implements the $\frac{1}{v^g}\frac{1}{p}\frac{dp}{dt}$ term in the shape equation.  The auxiliary kernel alters how the precursors are solved, the method is outlined by \sct{sect:prec}.  The executioner is where the PRKE are evaluated after shape solves are performed.  The postprocessors are used to evaluate the parameters for the PRKE and a user object is used to pass them to the executioner.  IQS's implementation the action system made sure that all the correct kernels and postprocessors are executed when the IQS executioner is specified in the input. \\

@@

@@

%Equations (\ref{eq:p}) and (\ref{eq:c}) are a formulation of the point reactor kinetics equation (PRKE), but the parameters (Equations (\ref{eq:rmb})-(\ref{eq:l})) are dependent on the shape.  If the assumption is made that the shape is time-independent, the shape is computed once at the first time step and used for the PRKE parameter evaluation at all other steps.  However, if the shape is dependent on time, the shape needs to be computed in transient using equation (\ref{eq:shape}) and (\ref{eq:prec}) in order retain accuracy.  Equations (\ref{eq:shape2}) and (\ref{eq:preq2}) shows the usual form of the shape and precursor equations with amplitude put on the right hand side.  Equation (\ref{eq:shape2}) is very similar to the multigroup flux equation (\ref{eq:flux}), except the removal cross-section term is augmented by a $\frac{1}{v^g}\frac{1}{p}\frac{dp}{dt}$ term and the precursor contribution has a $\frac{1}{p}$ multiplier.  Equation (\ref{eq:preq2}) is very similar to the normal precursor equation (\ref{eq:precursor}), except the fission source term is multiplied by $p$.  These differences are crucial for IQS implementation in \rattlesnake.

>>>>

% input stuff
\lstset{%
%  language=C++,
  showstringspaces=false,
% basicstyle=\scriptsize\ttfamily,
  basicstyle=\footnotesize\ttfamily,
  commentstyle=\color{inl@green},
%  keywordstyle=\bfseries,
%  escapeinside=$$
%  framesep=0pt,
%  rulesep=0pt,
  frame=single
}
% Make nice big tildes inside lstlistings instead of
% tiny ones.
\lstset{
    literate={~} {$\sim$}{1}
}

@@

\rattlesnake is a MOOSE-based application developed INL specific to solving radiation transport problems with multiphysics capabilities.  MOOSE is a finite-element based, multiphysics framework that gives the general architecture for the development of physics application like Rattlesnake.   At the heart of \rattlesnake is the action system, which provides a means to consolidate the MOOSE input syntax (which can be quite larger for multigroup transport simulations), so that a user does not have to define every kernel, variable, etc., used in the problem. Rather, the user inputs an equation description 
(e.g., Diffusion, S$_n$, P$_n$, etc.) and a solution method (e.g., SAAF, LS, CFEM, DFEM, etc.), and the action system will incorporate all the necessary physics involved (kernels, boundary conditions, postprocessor, etc.).

@@

The input file for IQS is very similar to the current transient diffusion input file.  The IQS input has a different executioner type and parameters.  The executioner type is simply IQS and input parameters include the number of micro time steps per macro step, the IQS error tolerance, and the initial power.  The \rattlesnake transient action system currently requires a multi-app and transfer to compute and pass the initial $\phi$ and $\keff$, which is present in the transient input deck.  However, IQS also requires an initial evaluation of the adjoint flux, for the weighting function.  So another input file and multi-app transfer was made for the adjoint calculation.  \vspace{3mm}

@@

\fig{fig:TWIGL_power} shows the IQS  solution as compared with the Brute Force solution.  These plots show that IQS is consistent in more complex, higher dimensional problems in Rattlesnake.  Finally, \fig{fig:TWIGL_conv} plots the error convergence of IQS and the Brute Force methods.  The curves show the impressive convergence of IQS for the highly transient TWIGL example.  \tbl{tab:TWIGLdt2} and \fig{fig:TWIGL_power_dt2} show the results for time adaptation.  The results show that both IQS methods perform exceptionally well compared to brute force.  It also shows that traditional IQS performed better with large $e_{tol}$, while IQS P-C was better with smaller $e_{tol}$.

@@

The \fig{fig:LRA_plots} plots show the IQS  solution as compared with the Brute Force solution.    These plots show that IQS is consistent with Doppler feedback problems in Rattlesnake.  \fig{fig:lra_conv} plots the error convergence of IQS and the Brute Force methods. \tbl{tab:LRAdt2} shows the time adaptation performance of brute force and IQS P-C.  These results show that IQS, again, performs much better than the brute force method by taking significantly fewer diffusion solves while retaining similar accuracy.

@@

The purpose of this report is to briefly review the improved quasi-static method for neutronic transient simulations, to describe its implementation in Rattlesnake, the radiation transport application based on MOOSE, and to show its effectiveness using several test cases.  
%The derivation of IQS was done with multigroup diffusion, but it was seen in \sct{sect:transport} that application to transport is relatively simple.  
Two IQS implementations are available in Rattlesnake.  The first one is the traditional IQS approach, where the shape equation and PRKE are solved as a coupled nonlinear system with Picard iterations.  The second approach, IQS P-C, is a predictor-corrector linearization, where the flux is solved and then rescaled using the amplitude obtained through the PRKE
(Point Reactor Kinetic Equations). \vspace{2mm}

IQS's implementation in \rattlesnake involves a new kernel, an auxiliary kernel, a new executioner derived from the current transient, four postprocessors, as well as alterations to the action system.  The new kernel implements the $\frac{1}{v^g}\frac{1}{p}\frac{dp}{dt}$ term in the shape equation.  The auxiliary kernel alters how the precursors are solved, the method is outlined by \sct{sect:prec}.  The executioner is where the PRKE are evaluated after shape solves are performed.  The postprocessors are used to evaluate the parameters for the PRKE and a user object is used to pass them to the executioner.  IQS's implementation the action system made sure that all the correct kernels and postprocessors are executed when the IQS executioner is specified in the input. \vspace{2mm}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Future Work}

Now that we are confident in the transient modeling capabilities with IQS, our next step in this research will be to compare results with a direct Rattlesnake solution for TREAT.  INL has performed a number of transient tests for various TREAT configurations with thermal feedback.  Use of the MAMMOTH reactor physics app will allow coupling of Rattlesnake (including the new IQS routines) with BISON.  In this application, we will need to investigate the temporal coupling of BISON with the IQS time stepping process.  It may or may not be necessary to couple each micro time step to BISON updates, but further research in this area is needed.  Another interest to improve IQS's capability is to develop a smarter time adaptation scheme.  The adaptation would entail predicting the perturbation of the shape based on transient changes in material properties.  The ability to update PRKE parameters within macro steps may also be a valuable improvement.

@@

%Equations (\ref{eq:p}) and (\ref{eq:c}) are a formulation of the point reactor kinetics equation (PRKE), but the parameters (Equations (\ref{eq:rmb})-(\ref{eq:l})) are dependent on the shape.  If the assumption is made that the shape is time-independent, the shape is computed once at the first time step and used for the PRKE parameter evaluation at all other steps.  However, if the shape is dependent on time, the shape needs to be computed in transient using equation (\ref{eq:shape}) and (\ref{eq:prec}) in order retain accuracy.  Equations (\ref{eq:shape2}) and (\ref{eq:preq2}) shows the usual form of the shape and precursor equations with amplitude put on the right hand side.  Equation (\ref{eq:shape2}) is very similar to the multigroup flux equation (\ref{eq:flux}), except the removal cross-section term is augmented by a $\frac{1}{v^g}\frac{1}{p}\frac{dp}{dt}$ term and the precursor contribution has a $\frac{1}{p}$ multiplier.  Equation (\ref{eq:preq2}) is very similar to the normal precursor equation (\ref{eq:precursor}), except the fission source term is multiplied by $p$.  These differences are crucial for IQS implementation in Rattlesnake.
